Computational methods in landscape ecology

### 2024-08-29 Manuscript.tex
Updating software section and tiny cosmetic changes

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=25mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

Landscape simulators are typically used as null hypotheses, baselines,
or scenarios because they allow to control certain aspects of the
landscape (Li et al. 2004; Wang and Malanson 2008) and are generally
simpler than ecological simulations models (Hess et al. 2020). Landscape
simulators can be classified into two major categories, namely
pattern-based and process-based approaches {[}Pe'er et al. (2013);
Langhammer2019{]}.

@@

``Open science'' involves full transparency and data access at all
stages of the research processes and the integrative and
multi-disciplinary field of ecology can massively benefit from this
concept (Hampton et al. 2015). One key aspect of achieving open science
is to use open source software, which ensures transparency and
consequently increasing the (public) trust into the analyses (Powers and
Hampton 2019). Related to this, script-based workflows have the
advantage over workflows involving graphical user interfaces that the
former are shareable and reproducible (Hampton et al. 2015; Powers and
Hampton 2019). Thus, we will focus the following section on script-based
open-source software suitable for landscape ecology. Given the
increasing availability of data and complexity of methods, future
progress within the field of landscape ecology will depend on the
software and analysis tools available (Costanza et al. 2019)

The open-source and script-based R programming languages (R Core Team
2024) was first released in 1995 (Smith 2016) with a focus on
statistical computing and data visualization. However, due to its
extensions (also called packages), the language is popular in many
academic fields including ecology (Lai et al. 2019) and it is suitable
for spatial data analysis or landscape ecology (e.g., Bivand 2020;
Hesselbarth et al. 2021).

@@

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Li2004Use}
Li H, Wu J (2004) Use and misuse of landscape indices. Landscape ecology
19:389--399

\bibitem[\citeproctext]{ref-Li2004}
Li X, He HS, Wang X, et al (2004) Evaluating the effectiveness of
neutral landscape models to represent a real landscape. Landscape and
Urban Planning 69:137--148.
\url{https://doi.org/10.1016/j.landurbplan.2003.10.037}

@@

@@

@@

@@

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-Turner1989Landscape}
Turner MG (1989) Landscape ecology: The effect of pattern on process.
Annual Review of Ecology and Systematics 20:171--197.
\url{https://doi.org/10.1146/annurev.es.20.110189.001131}

@@

\bibitem[\citeproctext]{ref-Turner1989Effects}
Turner MG, O'Neill RV, Gardner RH, Milne BT (1989) Effects of changing
spatial scale on the analysis of landscape pattern. Landscape ecology
3:153--162

@@

@@

\bibitem[\citeproctext]{ref-Wu2002Key}
Wu J, Hobbs R (2002) Key issues and research priorities in landscape
ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
\url{https://doi.org/10.1023/A:1020561630963}

\bibitem[\citeproctext]{ref-Wu2002Empirical}
Wu J, Shen W, Sun W, Tueller PT (2002) Empirical patterns of the effects
of changing scale on landscape metrics. Landscape ecology 17:761--782

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=25mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

Landscape simulators are typically used as null hypotheses, baselines,
or scenarios because they allow to control certain aspects of the
landscape (Li et al. 2004; Wang and Malanson 2008) and are generally
simpler than ecological simulations models (Hess et al. 2020). Landscape
simulators can be classified into two major categories, namely
pattern-based and process-based approaches (Pe'er et al. 2013;
Langhammer et al. 2019).

@@

``Open science'' involves full transparency and data access at all
stages of the research processes and integrative and multi-disciplinary
fields such as ecology can massively benefit from this concept (Hampton
et al. 2015). Using open source software is one key aspect of open
science because it ensures transparency and consequently increases trust
in the analyses (Powers and Hampton 2019). Given the increasing data
availability and method complexity, future progress within the research
field will highly depend on available software and analysis tools
(Costanza et al. 2019). In this context, script-based workflows have an
advantage over workflows involving graphical user interfaces because the
former are more shareable and reproducible (Hampton et al. 2015; Powers
and Hampton 2019). Thus, we will focus the following section on
script-based open-source software suitable for landscape ecology.

The open-source, script-based R programming languages (R Core Team 2024)
was first released in 1995 (Smith 2016) with a focus on statistical
computing and data visualization. However, due to its extensions (also
called packages), the language is popular in many academic fields,
including ecology (Lai et al. 2019; Lortie et al. 2020), and it is
suitable for spatial data analysis and landscape ecology (e.g., Bivand
2020; Hesselbarth et al. 2021). Fundamental packages for spatial data
analysis include \emph{sf} for simple-features vector data (Pebesma
2018), \emph{terra} for raster data (Hijmans 2024), \emph{stars} for
spatiotemporal arrays, raster, and vector data cubes (Pebesma and Bivand
2023), and \emph{spatstat} for point pattern data (Baddeley et al.
2015). Several packages allow to create maps, such as \emph{tmap}
(Tennekes 2018) or \emph{ggplot2} (Wickham 2016). For a comprehensive
overview of packages related to geographical spatial data, please see
the CRAN Task View (Bivand and Nowosad 2024).

R packages developed specifically for spatial pattern analysis include,
besides others, \emph{landscapemetrics} (Hesselbarth et al. 2019),
\emph{vectormetrics} {[}still being developed; Matuszek et al.
(2024){]}, \emph{motif} (Nowosad 2021), \emph{rasterdiv} (Rocchini et
al. 2021), or \emph{geodiv} (Smith et al. 2021). Packages with a focus
on scale include e.g., \emph{multilandr} (Huais 2024) or
\emph{grainchanger} (Graham 2024). Connectivity analysis can be
conducted using packages including \emph{lconnect} (Mestre and Silva
2023), or \emph{grainscape} (Chubaty et al. 2020). Furthermore, neutral
landscapes can be simulated using \emph{NLMR} (Sciaini et al. 2018), and
the \emph{graph4lg} (Savary et al. 2021) provides functions for
landscape genetics.

@@

\bibitem[\citeproctext]{ref-Baddeley2015}
Baddeley A, Rubak E, Turner R (2015) Spatial point patterns: Methodology
and applications with r. Chapman \& Hall/CRC, London, UK

@@

\bibitem[\citeproctext]{ref-Bivand2024}
Bivand RS, Nowosad J (2024) CRAN task view: Analysis of spatial data.
{https://CRAN.R-project.org/view=Spatial}

@@

\bibitem[\citeproctext]{ref-Chubaty2020a}
Chubaty AM, Galpern P, Doctolero SC (2020) The r toolbox grainscape for
modelling and visualizing landscape connectivity using spatially
explicit networks. Methods in Ecology and Evolution 11:591--595.
\url{https://doi.org/10.1111/2041-210X.13350}

@@

\bibitem[\citeproctext]{ref-Graham2024}
Graham L (2024) Grainchanger: Moving-window and direct data aggregation.
{https://ropensci.github.io/grainchanger/}

@@

\bibitem[\citeproctext]{ref-Hijmans2024}
Hijmans RJ (2024) Terra: Spatial data analysis. R package version
1.7-78. {\(<\)}{https://CRAN.R-project.org/package=terra\(>\)}

@@

\bibitem[\citeproctext]{ref-Li2004b}
Li H, Wu J (2004) Use and misuse of landscape indices. Landscape ecology
19:389--399

\bibitem[\citeproctext]{ref-Li2004a}
Li X, He HS, Wang X, et al (2004) Evaluating the effectiveness of
neutral landscape models to represent a real landscape. Landscape and
Urban Planning 69:137--148.
\url{https://doi.org/10.1016/j.landurbplan.2003.10.037}

\bibitem[\citeproctext]{ref-Lortie2020}
Lortie CJ, Braun J, Filazzola A, Miguel F (2020) A checklist for
choosing between r packages in ecology and evolution. Ecology and
Evolution 10:1098--1105. \url{https://doi.org/10.1002/ece3.5970}

@@

\bibitem[\citeproctext]{ref-Matuszek2024}
Matuszek T, Nowosad J, Sciaini M, et al (2024) Vectormetrics: Landscape
metrics for categorical map patterns in vector data.
{https://github.com/r-spatialecology/vectormetrics}

@@

\bibitem[\citeproctext]{ref-Mestre2023}
Mestre F, Silva B (2023) Lconnect r package: A versatile tool for
evaluating landscape connectivity and prioritizing habitat patches in
conservation research. Ecological Modelling 484:110489.
\url{https://doi.org/10.1016/j.ecolmodel.2023.110489}

@@

\bibitem[\citeproctext]{ref-Pebesma2018}
Pebesma E (2018) Simple features for r: Standardized support for spatial
vector data. The R Journal 10:439.
\url{https://doi.org/10.32614/RJ-2018-009}

\bibitem[\citeproctext]{ref-Pebesma2023}
Pebesma E, Bivand R (2023)
\href{https://doi.org/10.1201/9780429459016}{Spatial data science: With
applications in r}, 1st edn. {Chapman and Hall/CRC}, New York

@@

\bibitem[\citeproctext]{ref-Rocchini2021}
Rocchini D, Thouverai E, Marcantonio M, et al (2021) Rasterdiv---an
information theory tailored r package for measuring ecosystem
heterogeneity from space: To the origin and back. Methods in Ecology and
Evolution 12:1093--1102. \url{https://doi.org/10.1111/2041-210X.13583}

@@

\bibitem[\citeproctext]{ref-Savary2021}
Savary P, Foltête J-C, Moal H, et al (2021) graph4lg: A package for
constructing and analysing graphs for landscape genetics in r. Methods
in Ecology and Evolution 12:539--547.
\url{https://doi.org/10.1111/2041-210X.13530}

@@

\bibitem[\citeproctext]{ref-Sciaini2018}
Sciaini M, Fritsch M, Scherer C, Simpkins CE (2018) NLMR and
landscapetools: An integrated environment for simulating and modifying
neutral landscape models in r. Methods in Ecology and Evolution
9:2240--2248. \url{https://doi.org/10.1111/2041-210X.13076}

@@

\bibitem[\citeproctext]{ref-Smith2021b}
Smith AC, Dahlin KM, Record S, et al (2021) The {\textsc{geodiv r}}
package: Tools for calculating gradient surface metrics. Methods in
Ecology and Evolution 12:2094--2100.
\url{https://doi.org/10.1111/2041-210X.13677}

@@

\bibitem[\citeproctext]{ref-Tennekes2018}
Tennekes M (2018) Tmap: Thematic maps in r. Journal of Statistical
Software 84: \url{https://doi.org/10.18637/jss.v084.i06}

\bibitem[\citeproctext]{ref-Turner1989a}
Turner MG (1989) Landscape ecology: The effect of pattern on process.
Annual Review of Ecology and Systematics 20:171--197.
\url{https://doi.org/10.1146/annurev.es.20.110189.001131}

@@

\bibitem[\citeproctext]{ref-Turner1989b}
Turner MG, O'Neill RV, Gardner RH, Milne BT (1989) Effects of changing
spatial scale on the analysis of landscape pattern. Landscape ecology
3:153--162

@@

\bibitem[\citeproctext]{ref-Wickham2016}
Wickham H (2016) ggplot2: Elegant graphics for data analysis. Springer
New York, New York, USA

@@

\bibitem[\citeproctext]{ref-Wu2002b}
Wu J, Hobbs R (2002) Key issues and research priorities in landscape
ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
\url{https://doi.org/10.1023/A:1020561630963}

\bibitem[\citeproctext]{ref-Wu2002a}
Wu J, Shen W, Sun W, Tueller PT (2002) Empirical patterns of the effects
of changing scale on landscape metrics. Landscape ecology 17:761--782

### 2024-08-29 Sections/01-introduction.qmd
Updating software section and tiny cosmetic changes

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

>>>>

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989a; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002b; @Wu2013].

### 2024-08-29 Sections/02-spatial-patterns.qmd
Updating software section and tiny cosmetic changes

<!-- Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. -->
Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003].
However, spatial patterns are not only a result of processes, but could also be a driver of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989Landscape].

@@

<!-- https://doi.org/10.1371/journal.pone.0304215 -->
Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004Use; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation. Let us go through them in a different order: -->
In line with Dramstad [@Dramstad2009], we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation.

@@

<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989Effects; @Saura2001; @Wu2002Empirical; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches, resulting in unreliable landscape metric computations.
In order to address scale issues and advance the science of scaling in landscape ecology, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
As suggested by Frazier [@Frazier2023], this can be done following three key steps, namely by considering scope guidelines when designing experiments, explicitly reporting grain and extent so that scope can be calculated allowing quantitative comparability with other studes and undertaking comparisons and replications based on scope.
The latter is showcased in a proof of concept example experiment, where landscapes from the same ecoregion are sampled using different grain and extent and then the distributions of the computed landscape metrics are visualized through ridgeline plots colored by scope [@Frazier2023].

>>>>

Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003].
However, spatial patterns are not only a result of processes, but could also be a driver of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989a].

@@

<!-- https://doi.org/10.1371/journal.pone.0304215 -->
Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation. Let us go through them in a different order: -->
In line with Dramstad [@Dramstad2009], we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation.

@@

<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches, resulting in unreliable landscape metric computations.
In order to address scale issues and advance the science of scaling in landscape ecology, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
As suggested by Frazier [@Frazier2023], this can be done following three key steps, namely by considering scope guidelines when designing experiments, explicitly reporting grain and extent so that scope can be calculated allowing quantitative comparability with other studes and undertaking comparisons and replications based on scope.
The latter is showcased in a proof of concept example experiment, where landscapes from the same ecoregion are sampled using different grain and extent and then the distributions of the computed landscape metrics are visualized through ridgeline plots colored by scope [@Frazier2023].

### 2024-08-29 Sections/05-simulations.qmd
Updating software section and tiny cosmetic changes

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; Langhammer2019].

>>>>

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004a; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; @Langhammer2019].

### 2024-08-29 Sections/08-software.qmd
Updating software section and tiny cosmetic changes

<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->

"Open science" involves full transparency and data access at all stages of the research processes and the integrative and multi-disciplinary field of ecology can massively benefit from this concept [@Hampton2015].
One key aspect of achieving open science is to use open source software, which ensures transparency and consequently increasing the (public) trust into the analyses [@Powers2019].
Related to this, script-based workflows have the advantage over workflows involving graphical user interfaces that the former are shareable and reproducible [@Hampton2015; @Powers2019].
Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.
Given the increasing availability of data and complexity of methods, future progress within the field of landscape ecology will depend on the software and analysis tools available [@Costanza2019]

<!-- R: Hesselbarth -->
The open-source and script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
However, due to its extensions (also called packages), the language is popular in many academic fields including ecology [@Lai2019] and it is suitable for spatial data analysis or landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].

<!-- 
landscapemetrics
geodiv
vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
bespatial
LandComp
motif
?GLCMTextures
?rasterdiv
?terra 
-->

@@

<!--Others: -->

<!-- 
GuidosToolbox
?GRASS GIS
?QGIS: LecoS plugin 
-->

>>>>

"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields such as ecology can massively benefit from this concept [@Hampton2015].
Using open source software is one key aspect of open science because it ensures transparency and consequently increases trust in the analyses [@Powers2019].
Given the increasing data availability and method complexity, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019].
In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015; @Powers2019].
Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.

<!-- R: Hesselbarth -->
The open-source, script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
However, due to its extensions (also called packages), the language is popular in many academic fields, including ecology [@Lai2019; @Lortie2020], and it is suitable for spatial data analysis and landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].
Fundamental packages for spatial data analysis include *sf* for simple-features vector data [@Pebesma2018], *terra* for raster data [@Hijmans2024], *stars* for spatiotemporal arrays, raster, and vector data cubes [@Pebesma2023], and *spatstat* for point pattern data [@Baddeley2015].
Several packages allow to create maps, such as *tmap* [@Tennekes2018] or *ggplot2* [@Wickham2016]. 
For a comprehensive overview of packages related to geographical spatial data, please see the CRAN Task View [@Bivand2024].

R packages developed specifically for spatial pattern analysis include, besides others, *landscapemetrics* [@Hesselbarth2019], *vectormetrics* [still being developed; @Matuszek2024], *motif* [@Nowosad2021a], *rasterdiv* [@Rocchini2021], or *geodiv* [@Smith2021b]. 
<!-- BESSpatial: https://github.com/Nowosad/bespatial/ -->
Packages with a focus on scale include e.g., *multilandr* [@Huais2024] or *grainchanger* [@Graham2024].
Connectivity analysis can be conducted using packages including *lconnect* [@Mestre2023], or *grainscape* [@Chubaty2020a].
Furthermore, neutral landscapes can be simulated using *NLMR* [@Sciaini2018], and the *graph4lg* [@Savary2021] provides functions for landscape genetics.

@@
